Supplementary Information

### 2021-02-12 doc/manuscript/MEE_manuscript.Rmd
cleanup

Tree rings provide a uniquely valuable long-term record for understanding how climate and other environmental drivers shape forest productivity.
However, traditional dendrochronology methods aggregate growth records of multiple trees--often limited to dominant canopy individuals--into residual chronologies, and therefore cannot simultaneously account for the effects of climate, tree size, and slowly changing environmental drivers.
This has limited the potential to use tree-rings to understand forest productivity, its climate sensitivity, and its global change responses.
Here, we develop a new method to simultaneously model non-linear effects of objectively determined principle climate drivers, reconstructed tree diameter ($DBH$), and year in generalized least squares models that account for the temporal autocorrelation inherent to each individual tree's growth.
We apply this method to tree-ring data from `r n_trees_all` trees representing `r n_species` species at ten globally distributed sites.
Our analysis identified similar climate drivers to those obtained via traditional methods. 
Growth responses were predominantly positive to precipitation and negative to temperature, with both included in 80% of top models, and with non-linear responses prevalent (*>50%* of relationships).
Growth rates--expressed as ring widths, basal area increments, or biomass increments--varied non-linearly with $DBH$.
Interactions between $DBH$ and climate were common (*43%* of cases tested). 
Accounting for $DBH$, growth rate varied directionally over time in most species-- declining in *90%*.
These trends were largely attributable to successional stand dynamics as cohorts and stands age, which remain challenging to parse from global change drivers.
As a rigorous analytical framework for statistically modeling tree growth responses to the most important climate drivers, $DBH$, and year in nonlinear models, our method provides a parsimonious approach for characterizing multiple interacting drivers of tree growth, opening the door for novel analyses using tree-rings to improve our understanding of forest responses to global change.

@@

Tree rings provide a long-term record of annual growth increments that is invaluable for understanding forests in an era of global change [@amoroso_dendroecology_2017; @fritts_dendroecology_1989]. 
Spanning time scales of decades to centuries or even millennia, they provide by far the most robust method for characterization of the interannual climate sensitivity of tree growth [@fritts_tree_1976; @braker_measuring_2002] and how it is changing [@sniderhan_growth_2016; @maxwell_declining_2016; @babst_twentieth_2019; @wilmking_global_2020]. 
Combined with forest censuses, they can be used to estimate forest woody productivity [$ANPP_{woody}$; @graumlich_long-term_1989; @davis_forest_2009; @dye_comparing_2016; @teets_linking_2018] and its climate sensitivity [@klesse_sampling_2018; @teets_quantifying_2018; @helcoski_growing_2019]. 
They also provide the long-term perspective critical to understanding how slowly changing environmental drivers including rising atmospheric carbon dioxide (CO~2~) and directional climate change are influencing tree growth and forest productivity [e.g., @levesque_water_2017; @mathias_disentangling_2018; @walker_integrating_2020].
This information is critical to predicting forest responses to climate change, and thereby reducing the enormous uncertainty surrounding future contributions of Earth's forests to the global carbon cycle [@friedlingstein_climatecarbon_2006].
Yet, collection and analysis of dendrochronological records has been traditionally been optimized to detect climate signals rather than to predict forest productivity, its climate sensitivity, and how it may be changing [@cherubini_potential_1998; @nehrbassahles_influence_2014; @klesse_sampling_2018;@babst_when_2018; @wilmking_global_2020].
As a result, prevailing approaches hold a number of limitations for using tree-rings to address pressing questions concerning forest productivity in the current era of rapid environmental change.

Dendrochronological methods to characterize the climate sensitivity of tree growth have been optimized to obtain the strongest possible climate signal for climate reconstructions, but face limitations when it comes to characterizing the climate sensitivity of individual trees or $ANPP_{woody}$.
Traditional methods begin by fitting a function (commonly a spline) to the growth record captured by each core, extracting residuals around the long-term trends, and then standardizing and averaging the residuals across cores to form a species-level ring-width index chronology [@fritts_tree_1976; @speer_fundamentals_2010].
Climate signals are then identified by examining month-by-month or seasonal correlations of the ring-width index chronology to one or more climate variables over the current year and, typically, previous year [e.g., @fritts_tree_1976; @meko_seascorr_2011; @zang_treeclim_2015].
Following identification of the top climate driver(s), statistical models describing their relationships to tree growth can be used for applications such as climate reconstruction [e.g., @buntgen_2500_2011] or projection of tree growth responses to climate change [e.g., @charney_observed_2016].
An important caveat for the latter, however, is that the slopes of correlations between climate variables and ring-width index chronologies are not identical to the mean sensitivity of individuals within the population, as the process of building species chronologies fundamentally alters and obfuscates individual-level responses [@pederson_framework_2020].
Although population-level climate responses have been approximated based on climate sensitives derived from species chronologies [e.g., @charney_observed_2016; @helcoski_growing_2019], the removal of individual-level variation prior to analysis of climate sensitivity limits potential for using species chronologies to characterize the climate sensitivity of $ANPP_{woody}$.
In particular, analysis of species-level chronologies does not allow characterization of known variation in climate sensitivity in relation to tree size, canopy position, or microhabitat [e.g., @bennett_larger_2015; @mcgregor_tree_2020; @rollinson_climate_nodate].

Dendrochronological studies most commonly focus on linear climate responses to single climate drivers and the settings in which these occur, thereby missing  nonlinearities and additive or interactive climate effects known to be widespread within forest settings [@wilmking_global_2020]. 
Nonlinearities in climate sensitivities of tree metabolism and growth have been observed across a wide range spatio-temporal scales.
Over time frames of seconds to days, photosynthesis and respiration display a unimodal relationships to temporal variation in temperature, typically peaking at temperatures reflective of the environment to which the plant is adapted and acclimated [@kumarathunge_acclimation_2019]. 
Across broad geographical climate gradients, annual forest productivity generally increases with temperature, precipitation, and potential evapotranspiration ($PET$)  up to a point, after which it plateaus or decreases [@sullivan_long-term_2020; @banbury_morgan_global_nodate]. 
Filling a critical gap between short-term physiological responses and the global gradients representing millennia of community assembly and species adaptation, the annual growth records of tree-rings capture tree growth responses to interannual climatic variation.
Yet, because non-linearities are very problematic for reconstructing climate variables [@esper_divergence_2009], systems exhibiting these are typically avoided--e.g., by sampling climate-limited forest boundaries, and traditional analysis methods are designed around first-order linear growth-climate relationships [@fritts_tree_1976 **(dendro coauthors, does this ref work?)**]. 
Dendrochronological studies allowing for nonlinear or threshold responses of tree growth to climate are less common [@cook_climate_1989; @woodhouse_artificial_1999; @tolwinski-ward_bayesian_2013; @tumajer_increasing_2017; @cavin_highest_2017; @ljungqvist_assessing_2020; @rollinson_climate_nodate], and we therefore know little about the nonlinearities in growth responses to interannual variation in climate that occur for trees within forest settings.
Furthermore, temperature and moisture are known to jointly shape tree growth [@beedlow_importance_2013; @foster_predicting_2016] and forest productivity [e.g., @alexander_relative_2018; @banbury_morgan_global_nodate], yet growth sensitivity to their additive or interactive effects, potentially operating over different time windows, is not commonly considered [but see @meko_seascorr_2011; @foster_predicting_2016; @sanchez-salguero_disentangling_2015].

Tree diameter at breast height ($DBH$) is closely correlated with numerous variables affecting tree growth rate [e.g., height, crown size and position, root mass; @enquist_global_2002] and therefore is one of the most important variables influencing growth [e.g., @muller-landau_testing_2006; @foster_predicting_2016] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020].
However, for dendrochronological studies aimed at deciphering climate signals, $DBH$ is not typically a variable of interest, and its influence is removed through detrending [@cook_calculating_1997].
Moreover, many studies constrain sampling to only larger size classes.
While convenient for identifying climate signals [@fritts_tree_1976], this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity [@babst_when_2018].
Although climate correlations can be transformed to climate sensitivity [*i.e.*, magnitude of response, *sensu* @charney_observed_2016] and scaled to characterize the climate sensitivity of $ANPP_{woody}$ based on the scaling of $RW$ with $DBH$ [@helcoski_growing_2019], they cannot be used to directly model known interactive effects of $DBH$ and climate on tree growth [@trouillier_size_2019].
For example, larger trees tend to be relatively more sensitive to drought [@bennett_larger_2015; @gillerot_tree_2020; @mcgregor_tree_2020], and responses to temperature can also vary with tree size [@rossi_age-dependent_2007; @rollinson_climate_nodate].
However, the removal of $DBH$ signals via detrending makes it impossible to account for such size differences in climate sensitivity in a systematic, integrative way.
To use tree-rings to predict tree growth, $ANPP_{woody}$, and forest dynamics, we need models that include $DBH$ [e.g., @evans_fusing_2017; @klesse_amplifying_2020] and its potential climate interactions [e.g., @rollinson_climate_nodate].

@@

###other:
n_pos_first_order_P_term=42 #(exception: TSCA)
n_pos_1st_P_term_and_neg_2nd = 32
percent_temperature_responses_1st_order_negative=round(13/38*100)
percent_temperature_responses_1st_order_pos_2nd_order_neg=18/38*100
```
Responses to precipitation amount ($PPT$) and frequency ($PDF$) were included in the best model for all but one species, and were predominantly positive (Fig. `r clim_sens`).
$PPT$ was selected over $PDF$ as the top variable in five of the eight sites with both variables available (but had no significant main effect at one of these, NIO), and was the only option at two sites (LT and CB).
Optimal time windows most commonly coincided at least partially with the months of most active growth in the current year (Fig. `r clim_sens`; Table `r site_details`):  wet seasons in the tropics (BCNM and HKK) or late spring/ early summer in the extratropics (n= 5 of 8).
In the tropics, the long time windows over which precipitation was influential (12 mo at BCNM, 9 mo at HKK) also included the majority (BCNM) or all (HKK) of the dry season.
At LDW, the most influential windows were limited to the current spring and early summer, whereas optimal windows at three sites extended back to the previous fall (LT, CB) or summer (SCBI).
Finally, at three sites (HF, ZOF, and SC), precipitation of the previous growing season was the most influential variable. 
Responses to $PPT$ or $PDF$ were predominantly positive, with positive first-order linear terms for all but one species (*Tsuga canadensis* at HF; Fig. `r clim_sens`).
Negative second-order terms were commonly included in the best model (`r n_pos_1st_P_term_and_neg_2nd` of `r n_pos_first_order_P_term` with positive first-order terms), generally resulting in a deceleration or decline at the highest levels of precipitation, but occasionally producing a unimodal (e.g., several species at SCBI) or predominantly negative response (e.g., BEAL at HF; Fig. `r clim_sens`).

@@

Finally, our analysis differed from conventional approaches in the use of GLS models to simultaneously quantify the effects of climate, $DBH$, and year, while accounting for idiosyncratic growth trends of individual trees through an autocorrelation structure [step 2; Fig. `r schematic`; see also @evans_fusing_2017; @rollinson_climate_nodate].
This approach allowed the consideration of additive and interactive effects of climate with variables that change over longer time frames--here, $DBH$ and year, although other slowly-changing drivers could also be used.
We found that interactions between climate variables and $DBH$ were common (`r round(n_DBHx_analyzed_sig/ n_DBHx_analyzed*100)`%;  Figs. `r interactions_fig`, `r RW_interactions_all`, `r BAI_interactions_all`).
The most coherent pattern observed in this analysis was a tendency for larger trees to be more sensitive to precipitation and high temperatures (Fig. `r interactions_fig`), consistent with widespread observations that larger trees are more sensitive to drought [REFS; @bennett_larger_2015; @hacket-pain_consistent_2016; @pretzsch_drought_2018; @gillerot_tree_2020; @mcgregor_tree_2020].
An analytical structure that can account for this and other such DBH-climate interactions [e.g., @rossi_age-dependent_2007; @rollinson_climate_nodate] will be critical to using tree-ring records to understand and forecast the effects of climate on tree growth and forest productivity.
We note that a modification of our analysis method (modeling $year \times$ climate interactions instead of $climate \times$ climate interactions) could potentially be used to account for directional changes in climate sensitivity, which have commonly been documented [e.g., @maxwell_declining_2016; @helcoski_growing_2019; @peltier_tree_2020; @zuidema_recent_2020].
In fact, a recent literature survey revealed that such non-stationarity in the climate sensitivity of tree growth occurs globally and in the majority of tested cases [@wilmking_global_2020].

@@

Growth rates changed directionally for the majority (`r percent_sig_neg_trend_of_analyzed`, depending on growth metric) of site-species combinations selected for analysis based on the sampling representation of $DBH$ by $year$ (Fig. `r year_fig`).
These results were similar across growth metrics (Figs. `r all_GLS`), and consistent with an independent analysis of growth-$DBH$ trends by decade (Figs. `r year_fig`, `r decadal_analysis_figs_all`), indicating that our approach is robust in its analytical structure.
Indeed, in a comparative analysis of several methods commonly used to detect growth trends, an approach parallel to that employed here (regional curve standardization) performed better at growth trend detection [@peters_detecting_2015] than two of the most commonly used methods for analyzing growth trends, conservative detrending and basal area correction [see @peters_detecting_2015], both of which would fail to capture the observed trends in $RW$ and $BAI$ with $DBH$ (Fig. `r size`).
For instance, based on our finding that $BAI$ tends to rise and fall with $DBH$, basal area correction would generally be biased towards finding positive trends for smaller trees and negative trends for larger trees.
However, our results remain subject to the potential influence of some potential sampling and survivorship biases that can influence overall trends in positive or negative directions [Fig. `r year_fig`; @brienen_detecting_2012; @groenendijk_no_201; @brienen_tree_2017; @nehrbassahles_influence_2014].
While our analysis was designed to avoid some of the most severe potential biases [@brienen_detecting_2012], it is difficult, if not impossible, to control for all potential demography and survivorship biases, or to design sampling in a way that ensures unbiased representation of a species' growth rate at all points in the history of a stand [@brienen_detecting_2012; @brienen_tree_2017; @bowman_detecting_2013].
The observed trends should therefore be interpreted with caution, and as representative of only the sampled trees, as opposed to all individuals of the species that existed throughout the time frame analyzed.
Within this context, signals of changing growth rate over time are attributable to some combination of stand dynamics (e.g., recruitment and succession, changing stand structure) and environmental drivers (e.g., climate drivers other than those selected by *climwin*, rising atmospheric CO~2~, deposition of SO~2~ and NO~x~).

>>>>

Tree rings provide a uniquely valuable long-term record for understanding how climate and other environmental drivers shape forest productivity.
However, traditional dendrochronology methods aggregate growth records of multiple trees--often limited to dominant canopy individuals--into residual chronologies, and therefore cannot simultaneously account for the effects of climate, tree size, and slowly changing environmental drivers.
This has limited the potential to use tree-rings to understand forest productivity, its climate sensitivity, and its global change responses.
Here, we develop a new method to simultaneously model non-linear effects of objectively determined principle climate drivers, reconstructed tree diameter ($DBH$), and year in generalized least squares models that account for the temporal autocorrelation inherent to each individual tree's growth.
We apply this method to tree-ring data from `r n_trees_all` trees representing `r n_species` species at ten globally distributed sites.
Our analysis identified similar climate drivers to those obtained via traditional methods. 
Growth responses were predominantly positive to precipitation and negative to temperature, with both included in 78% of top models, and with non-linear responses prevalent (*>50%* of relationships).
Growth rates--expressed as ring widths, basal area increments, or biomass increments--varied non-linearly with $DBH$.
Interactions between $DBH$ and climate were common (44% of cases tested). 
Accounting for $DBH$, growth rate varied directionally over time in most species-- declining in 90%.
These trends were largely attributable to successional stand dynamics as cohorts and stands age, which remain challenging to parse from global change drivers.
As a rigorous analytical framework for statistically modeling tree growth responses to the most important climate drivers, $DBH$, and year in nonlinear models, our method provides a parsimonious approach for characterizing multiple interacting drivers of tree growth, opening the door for novel analyses using tree-rings to improve our understanding of forest responses to global change.

@@

Tree rings provide a long-term record of annual growth increments that is invaluable for understanding forests in an era of global change [@amoroso_dendroecology_2017; @fritts_dendroecology_1989]. 
Spanning time scales of decades to centuries or even millennia, they provide by far the most robust method for characterization of the interannual climate sensitivity of tree growth [@fritts_tree_1976; @braker_measuring_2002] and how it is changing [@sniderhan_growth_2016; @maxwell_declining_2016; @babst_twentieth_2019; @wilmking_global_2020]. 
Combined with forest censuses, they can be used to estimate forest woody productivity [$ANPP_{woody}$; @graumlich_long-term_1989; @davis_forest_2009; @dye_comparing_2016; @teets_linking_2018] and its climate sensitivity [@klesse_sampling_2018; @teets_quantifying_2018; @helcoski_growing_2019]. 
They also provide the long-term perspective critical to understanding how slowly changing environmental drivers including rising atmospheric carbon dioxide (CO~2~) and directional climate change are influencing tree growth and forest productivity [e.g., @levesque_water_2017; @mathias_disentangling_2018; @walker_integrating_2020].
This information is critical to predicting forest responses to climate change, and thereby reducing the enormous uncertainty surrounding future contributions of Earth's forests to the global carbon cycle [@friedlingstein_climatecarbon_2006].
Yet, collection and analysis of dendrochronological records has traditionally been optimized to detect climate signals rather than to predict forest productivity, its climate sensitivity, and how it may be changing [@cherubini_potential_1998; @nehrbassahles_influence_2014; @klesse_sampling_2018;@babst_when_2018; @wilmking_global_2020].
As a result, prevailing approaches hold a number of limitations for using tree-rings to address pressing questions concerning forest productivity in the current era of rapid environmental change.

Dendrochronological methods to characterize the climate sensitivity of tree growth have been optimized to obtain the strongest possible climate signal for climate reconstructions, but face limitations when it comes to characterizing the climate sensitivity of individual trees or $ANPP_{woody}$.
Traditional methods begin by fitting a function (commonly a spline) to the growth record captured by each core, extracting residuals around the long-term trends, and then standardizing and averaging the residuals across cores to form a species-level ring-width index chronology [@fritts_tree_1976; @speer_fundamentals_2010].
Climate signals are then identified by examining month-by-month or seasonal correlations of the ring-width index chronology to one or more climate variables over the current year and, typically, previous year [e.g., @fritts_tree_1976; @meko_seascorr_2011; @zang_treeclim_2015].
Following identification of the top climate driver(s), statistical models describing their relationships to tree growth can be used for applications such as climate reconstruction [e.g., @buntgen_2500_2011] or projection of tree growth responses to climate change [e.g., @charney_observed_2016].
An important caveat for the latter, however, is that the slopes of correlations between climate variables and ring-width index chronologies are not identical to the mean sensitivity of individuals within the population, as the process of building species chronologies fundamentally alters and obfuscates individual-level responses [@pederson_framework_2020].
Although population-level climate responses have been approximated based on climate sensitives derived from species chronologies [e.g., @charney_observed_2016; @helcoski_growing_2019], the removal of individual-level variation prior to analysis of climate sensitivity limits potential for using species chronologies to characterize the climate sensitivity of $ANPP_{woody}$.
In particular, analysis of species-level chronologies does not allow characterization of known variation in climate sensitivity in relation to tree size, canopy position, or microhabitat [e.g., @bennett_larger_2015; @mcgregor_tree_2020; @rollinson_climate_2021].

Dendrochronological studies most commonly focus on linear climate responses to single climate drivers and the settings in which these occur, thereby missing  nonlinearities and additive or interactive climate effects known to be widespread within forest settings [@wilmking_global_2020]. 
Nonlinearities in climate sensitivities of tree metabolism and growth have been observed across a wide range spatio-temporal scales.
Over time frames of seconds to days, photosynthesis and respiration display a unimodal relationships to temporal variation in temperature, typically peaking at temperatures reflective of the environment to which the plant is adapted and acclimated [@kumarathunge_acclimation_2019]. 
Across broad geographical climate gradients, annual forest productivity generally increases with temperature, precipitation, and potential evapotranspiration ($PET$)  up to a point, after which it plateaus or decreases [@sullivan_long-term_2020; @banbury_morgan_global_nodate]. 
Filling a critical gap between short-term physiological responses and the global gradients representing millennia of community assembly and species adaptation, the annual growth records of tree-rings capture tree growth responses to interannual climatic variation.
Yet, because non-linearities are very problematic for reconstructing climate variables [@esper_divergence_2009], systems exhibiting these are typically avoided--e.g., by sampling climate-limited forest boundaries, and traditional analysis methods are designed around first-order linear growth-climate relationships [@fritts_tree_1976 **(dendro coauthors, does this ref work?)**]. 
Dendrochronological studies allowing for nonlinear or threshold responses of tree growth to climate are less common [@cook_climate_1989; @woodhouse_artificial_1999; @tolwinski-ward_bayesian_2013; @tumajer_increasing_2017; @cavin_highest_2017; @ljungqvist_assessing_2020; @rollinson_climate_2021], and we therefore know little about the nonlinearities in growth responses to interannual variation in climate that occur for trees within forest settings.
Furthermore, temperature and moisture are known to jointly shape tree growth [@beedlow_importance_2013; @foster_predicting_2016] and forest productivity [e.g., @alexander_relative_2018; @banbury_morgan_global_nodate], yet growth sensitivity to their additive or interactive effects, potentially operating over different time windows, is not commonly considered [but see @meko_seascorr_2011; @foster_predicting_2016; @sanchez-salguero_disentangling_2015].

Tree diameter at breast height ($DBH$) is closely correlated with numerous variables affecting tree growth rate [e.g., height, crown size and position, root mass; @enquist_global_2002] and therefore is one of the most important variables influencing growth [e.g., @muller-landau_testing_2006; @foster_predicting_2016] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020].
However, for dendrochronological studies aimed at deciphering climate signals, $DBH$ is not typically a variable of interest, and its influence is removed through detrending [@cook_calculating_1997].
Moreover, many studies constrain sampling to only larger size classes.
While convenient for identifying climate signals [@fritts_tree_1976], this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity [@babst_when_2018].
Although climate correlations can be transformed to climate sensitivity [*i.e.*, magnitude of response, *sensu* @charney_observed_2016] and scaled to characterize the climate sensitivity of $ANPP_{woody}$ based on the scaling of $RW$ with $DBH$ [@helcoski_growing_2019], they cannot be used to directly model known interactive effects of $DBH$ and climate on tree growth [@trouillier_size_2019].
For example, larger trees tend to be relatively more sensitive to drought [@bennett_larger_2015; @gillerot_tree_2020; @mcgregor_tree_2020], and responses to temperature can also vary with tree size [@rossi_age-dependent_2007; @rollinson_climate_2021].
However, the removal of $DBH$ signals via detrending makes it impossible to account for such size differences in climate sensitivity in a systematic, integrative way.
To use tree-rings to predict tree growth, $ANPP_{woody}$, and forest dynamics, we need models that include $DBH$ [e.g., @evans_fusing_2017; @klesse_amplifying_2020] and its potential climate interactions [e.g., @rollinson_climate_2021].

@@

###other:
n_pos_first_order_P_term=42 #(exception: TSCA)
n_pos_1st_P_term_and_neg_2nd = 32
percent_temperature_responses_1st_order_negative=round(13/38*100)
percent_temperature_responses_1st_order_pos_2nd_order_neg=18/38*100
```
Responses to precipitation amount ($PPT$) and frequency ($PDF$) were included in the best model for all but two species, and were predominantly positive (Fig. `r clim_sens`).
$PPT$ was selected over $PDF$ as the top variable in five of the eight sites with both variables available (but had no significant main effect at one of these, NIO), and was the only option at two sites (LT and CB).
Optimal time windows most commonly coincided at least partially with the months of most active growth in the current year (Fig. `r clim_sens`; Table `r site_details`):  wet seasons in the tropics (BCNM and HKK) or late spring/ early summer in the extratropics (n= 5 of 8).
In the tropics, the long time windows over which precipitation was influential (12 mo at BCNM, 9 mo at HKK) also included the majority (BCNM) or all (HKK) of the dry season.
At LDW, the most influential windows were limited to the current spring and early summer, whereas optimal windows at three sites extended back to the previous fall (LT, CB) or summer (SCBI).
Finally, at three sites (HF, ZOF, and SC), precipitation of the previous growing season was the most influential variable. 
Responses to $PPT$ or $PDF$ were predominantly positive, with positive first-order linear terms for all but one species (*Tsuga canadensis* at HF; Fig. `r clim_sens`).
Negative second-order terms were commonly included in the best model (`r n_pos_1st_P_term_and_neg_2nd` of `r n_pos_first_order_P_term` with positive first-order terms), generally resulting in a deceleration or decline at the highest levels of precipitation, but occasionally producing a unimodal (e.g., several species at SCBI) or predominantly negative response (e.g., BEAL at HF; Fig. `r clim_sens`).

@@

Finally, our analysis differed from conventional approaches in the use of GLS models to simultaneously quantify the effects of climate, $DBH$, and year, while accounting for idiosyncratic growth trends of individual trees through an autocorrelation structure [step 2; Fig. `r schematic`; see also @evans_fusing_2017; @rollinson_climate_2021].
This approach allowed the consideration of additive and interactive effects of climate with variables that change over longer time frames--here, $DBH$ and year, although other slowly-changing drivers could also be used.
We found that interactions between climate variables and $DBH$ were common (`r round(n_DBHx_analyzed_sig/ n_DBHx_analyzed*100)`%;  Figs. `r interactions_fig`, `r RW_interactions_all`, `r BAI_interactions_all`).
The most coherent pattern observed in this analysis was a tendency for larger trees to be more sensitive to precipitation and high temperatures (Fig. `r interactions_fig`), consistent with widespread observations that larger trees are more sensitive to drought [REFS; @bennett_larger_2015; @hacket-pain_consistent_2016; @pretzsch_drought_2018; @gillerot_tree_2020; @mcgregor_tree_2020].
An analytical structure that can account for this and other such DBH-climate interactions [e.g., @rossi_age-dependent_2007; @rollinson_climate_2021] will be critical to using tree-ring records to understand and forecast the effects of climate on tree growth and forest productivity.
We note that a modification of our analysis method (modeling $year \times$ climate interactions instead of $climate \times$ climate interactions) could potentially be used to account for directional changes in climate sensitivity, which have commonly been documented [e.g., @maxwell_declining_2016; @helcoski_growing_2019; @peltier_tree_2020; @zuidema_recent_2020].
In fact, a recent literature survey revealed that such non-stationarity in the climate sensitivity of tree growth occurs globally and in the majority of tested cases [@wilmking_global_2020].

@@

Growth rates changed directionally for the majority (`r percent_sig_neg_trend_of_analyzed`, depending on growth metric) of site-species combinations selected for analysis based on the sampling representation of $DBH$ by $year$ (Fig. `r year_fig`).
These results were similar across growth metrics (Figs. `r all_GLS`), and consistent with an independent analysis of growth-$DBH$ trends by decade (Figs. `r year_fig`, `r decadal_analysis_figs_all`), indicating that our approach is robust in its analytical structure.
Indeed, in a comparative analysis of several methods commonly used to detect growth trends, an approach parallel to that employed here (regional curve standardization) performed better at growth trend detection [@peters_detecting_2015] than two of the most commonly used methods for analyzing growth trends, conservative detrending and basal area correction [see @peters_detecting_2015], both of which would fail to capture the observed trends in $RW$ and $BAI$ with $DBH$ (Fig. `r size`).
For instance, based on our finding that $BAI$ tends to rise and fall with $DBH$, basal area correction would generally be biased towards finding positive trends for smaller trees and negative trends for larger trees.
However, our results remain subject to the potential influence of some potential sampling and survivorship biases that can influence overall trends in positive or negative directions [Fig. `r year_fig`; @brienen_detecting_2012; @groenendijk_no_2015; @brienen_tree_2017; @nehrbassahles_influence_2014].
While our analysis was designed to avoid some of the most severe potential biases [@brienen_detecting_2012], it is difficult, if not impossible, to control for all potential demography and survivorship biases, or to design sampling in a way that ensures unbiased representation of a species' growth rate at all points in the history of a stand [@brienen_detecting_2012; @brienen_tree_2017; @bowman_detecting_2013].
The observed trends should therefore be interpreted with caution, and as representative of only the sampled trees, as opposed to all individuals of the species that existed throughout the time frame analyzed.
Within this context, signals of changing growth rate over time are attributable to some combination of stand dynamics (e.g., recruitment and succession, changing stand structure) and environmental drivers (e.g., climate drivers other than those selected by *climwin*, rising atmospheric CO~2~, deposition of SO~2~ and NO~x~).
